Intervention Bandits

### 2015-03-13 causal-bandit/c-bandit.tex
adds work trying to figure out what happens to UCB regret if you initialize with bounds from observational data

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}
\usepackage{parskip}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}

@@

Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$. We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and then we observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.

@@

\eq {
P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
 &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
 &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
 &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
}

@@

Let $\hat{\mu}_a$ be an empirical estimator for $P(Y|do(X_i=j))$ obtained via marginalization over $X_a$.

\eq {
\hat{\mu}_a &=
\begin{cases}
\frac{m_{a,1}}{n_{a,1}}p_a+\frac{m_{a,0}}{n_{a,0}}(1-p_a) & \text{if } a \neq{i} \\
\frac{m_{i,j}}{n_{i,j}} & \text{if } a=i \\
\end{cases}\\
\text{where:}\\
m_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b, Y = 1}_s \\
n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s \\
}

@@

\eq {
\hat{\mu} &= \sum^{K}_{a=1}w_a\hat{\mu}_a=w_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} w_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
}

If $p$ is not known, these expression are unchanged except that $p_a$ is replaced with $\hat{p}_a$

\eq{
\hat{p}_a = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
}
\subsection{Observe then exploit}

@@

\section{Observe until number of plausibly optional arms $< \alpha$}

@@

\subsection{Regret during observe phase}

@@

\subsection{Regret during UCB stage}

@@

\subsection{General rough thoughts}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{tikz} 
\usetikzlibrary{arrows,positioning} 
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
%\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=1em,
           minimum height=1em,
           font=\footnotesize,
           text centered,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

@@

Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$, figure . We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.

@@

\eqn {
P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
 &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
 &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
 \label{eqn:marginalizedEstimator}
 &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
}

@@

Considering equation \ref{eqn:marginalizedEstimator}, a natural way to construct an empirical estimator,  $\hat{\mu}_a$, for $P(Y|do(X_i=j))$ utilizing the information obtained by intervening on variable $a$ is:

\eqn {
\hat{\mu}_a &=
\begin{cases}
\label{eqn:marginalizedEstimator2}
\frac{m_{a,1}}{n_{a,1}}p_a+\frac{m_{a,0}}{n_{a,0}}(1-p_a) & \text{if } a \neq{i} \\
\frac{m_{i,j}}{n_{i,j}} & \text{if } a=i \\
\end{cases}\\
}
where $\frac{m_{a,b}}{n_{a,b}}$ is an empirical estimator for $P(Y|X_i = j,do(X_a = b))$,
\eqn{
m_{a,b}   &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b, Y = 1}_s \\
n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s 
}

@@

\eqn {
\label{eqn:pooledEstimators}
\hat{\mu} &= \sum^{K}_{a=1}w_a\hat{\mu}_a=w_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} w_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
}

If $p$ is not known $p_a$ could be replaced with its empirical estimate $\hat{p}_a$

\eqn{
\hat{p}_a = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
}

Now I want to select weights $w$ to minimize the variance in $\hat{\mu}$ and then get some kind of high probability bound for it. However, if I'm not very careful about how I pool things in equation \ref{eqn:pooledEstimators}, the estimator could get very biased, ie estimators $\hat{\mu_a}$ where I have data only for one state of the variable $a$ are not very informative, so should tend to get low weight, but if we choose to sample the second state based on previous payoffs - then we could end up assigning high weights to those estimators that had fluctuated above the true value...

This complicates getting correct uncertainty bounds for a UCB type algorithm/proof.

Consider, what exactly you mean by not unbiased ...

\section{Importance Sampling Approach}

Consider down weighting estimators proportional to the number of times we played them

\section{Observe then exploit}

\subsection{Basic idea}

@@

\subsection{Alternate Observe and Exploit With Some Decay}
Keep track of two separate bounds and always choose the narrower of the two.

\subsection{Observe until number of plausibly optional arms $< \alpha$}

@@

\subsubsection{Regret during observe phase}

@@

Clearly, this is not quite the right criteria to decide when to switch ...
\subsubsection{Regret during UCB stage}

@@

Suppose we have $O_i$ observations for arm $i$. We will now begin doing standard UCB - storing a separate set of bounds, only each time we choose an action we will pick the narrower of the observational/interventional confidence bounds.

\eqn{
\hat{\epsilon}_{it} &= \min \left( \sqrt{\frac{\alpha \log(t)}{2 T_{it}}},\sqrt{\frac{log(1/\delta)}{2 O_i}} \right)\\
\label{eqn:combinedBoundsEpsilon}
& = \sqrt{\frac{\alpha \log(t)}{2 \max(T_{it},O_i)}} \;\; \text{if we let $\delta = t^{-\alpha}$}
}

When running UCB, in order to select a non-optimal arm $i$ at timestep $t$, at least one of the following statements must be true:

\begin{enumerate}
\item $\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \mu_i + \Delta_i$
\item $\hat{\mu}_{it} + \hat{\epsilon}_{it} \geq \mu_i + 2\hat{\epsilon}_{it}$
\item $\Delta_i < 2 \hat{\epsilon}_{it}$
\end{enumerate}

Given our expression for the confidence bound in equation \ref{eqn:combinedBoundsEpsilon} we can write statement (3) as:
\eqn{
\max(T_{it},O_i) < \frac{2\alpha \log(t)}{\Delta_i^2} := \beta
}

If $O_i \geq \beta$, statement (3) is guaranteed to be false for all $t$ and we can bound the expected number of times we select $i$ with how many times we expect the bounds given by (1) and (2) to be true.

For arms $i$ such that $O_i \geq \beta$

\eqn{
&P(\text{1 is true}) < t^{1-\alpha}\\
&P(\text{2 is true}) < t^{1-\alpha}\\
& \implies P(I_t = i) < 2t^{1-\alpha}\\
& \implies R_n(i) \leq \frac{\Delta_i \alpha}{\alpha -2}
}

For all other arms with $\Delta_i > 0$
\eqn{
R_n(i) \leq  \frac{2\alpha \log(n)}{\Delta_i} + \frac{\Delta_i \alpha}{\alpha -2} 
}

Assume we observed for $h$ steps. The expected regret from this period is just $h E[\Delta_i]$. Let $T = h+n$ be the total time.

\eqn{
R_T < \sum_{i=1}^K \Delta_i \left(\frac{h}{K} + \frac{\alpha}{\alpha-2}\right) + \sum_{i:\Delta_i > 0}P(O_i \leq \beta)\frac{2\alpha \log(n)}{\Delta_i}
}

Using Hoeffding's Inequality:

\eqn {
P(O_i \leq \beta) \leq \begin{cases}
2 e^{-2h(\frac{1}{2}-\frac{\beta}{h})^2} & \text{if $h > 2\beta$} \\
1 & \text{otherwise}
\end{cases}
}

Its tricky - selecting the $h$ to switch at based on the inequality can certainly yield a non-optimal answer. Maybe I should figure this out via simulation?? But some theory to tell me what to simulate would certainly help.

\pagebreak

\subsubsection{General rough thoughts}
